### 2021-04-25 submission/manuscript.Rmd
Add references

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility. Our group has shown that when *C. difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C. difficile* negative with non-diarrheal consistency [@Schubert2015]. These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* (Ref). However, whether laxative results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization. For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure. Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C. difficile* spores and toxins from the intestine [@Liacouras1996].

In the past, our group has used C57BL6 mice to characterize how antibiotics including clindamycin disrupt the microbiota and influence *C. difficile* susceptibility and clearance [ref]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile*, these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance that should be addressed to better inform how we think about laxatives in the context of CDIs. Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in more severe CDI and sustained *C. difficile* colonization, and if PEG treatment post-CDI can promote *C. difficile* clearance.

@@

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy (ref), we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

@@

Next, we examined CDI severity by evaluating cecum and colon histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-infection (Fig. 3D), the timepoint typically examined in *C. difficile* 630 challenged mice (ref). We also looked at 6 days post-infection because that was when we started to see a large difference in C*. difficile* colonization levels between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is 12 (Fig. 3E). Thus, although PEG treatment had a profound impact on the mucosal microbiota, the impact of *C. difficile* 630 infection on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

* Discuss what these findings might mean for human patients (Fig. 8B)
  + What's known regarding laxatives and susceptibility to CDIs
  + Relevance to human FMTs? Unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs.

@@

## Materials and Methods

>>>>

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility. Our group has shown that when *C. difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C. difficile* negative with non-diarrheal consistency [@Schubert2014]. These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. However, whether laxative results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization. For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure. Many of the FMT studies to date rationalize the use of laxatives [@vanNood2013; @Razik2017; @Postigo2012] based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C. difficile* spores and toxins from the intestine [@Liacouras1996].

In the past, our group has used C57BL6 mice to characterize how antibiotics including clindamycin disrupt the microbiota and influence *C. difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance that should be addressed to better inform how we think about laxatives in the context of CDIs. Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in more severe CDI and sustained *C. difficile* colonization, and if PEG treatment post-CDI can promote *C. difficile* clearance.

@@

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy [@Nguyen2015], we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

@@

Next, we examined CDI severity by evaluating cecum and colon histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-infection (Fig. 3D), the timepoint typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-infection because that was when we started to see a large difference in C*. difficile* colonization levels between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is 12 (Fig. 3E). Thus, although PEG treatment had a profound impact on the mucosal microbiota, the impact of *C. difficile* 630 infection on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

* Discuss what these findings might mean for human patients (Fig. 8B)
  + What's known regarding laxatives and susceptibility to CDIs
  + Clinical trial of PEG, results never posted [@Dieterle2018]
  + Relevance to human FMTs? Unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs.

@@

## Materials and Methods
* Histopathology [@Theriot2011]
